---
title: "Relative Abundance Analysis from DADA2 Output"
author: "Jonas Ohlsson"
date: today
format:
  html:
    toc: true
    code-fold: false
    fig-width: 10
    fig-height: 7
execute:
  warning: true
  message: true
params:
  input_dir: "/Users/j/Documents/Gasfermentering/RISE/16S data"
  taxonomic_level: "genus"
  # color the bubble plot according to this taxonomic rank
  bubble_plot_taxonomic_level_color: "phylum"
  min_abundance: 0.025
  # do we want to agglomerate taxa at the specified level?
  agglomerate: false
  # include title and subtitle in plot?
  plot_titles: false
  # should we rename e.g., Methanobacteriales_2 to Methanobacteriales (2)?
  clean_names: false
  # file with additional sample metadata
  #   structure of the file:
  #   sample name in first column, metadata in other columns
  sample_data_file: "sample_data.xlsx"
  sample_type_column: "Substrate" # column containing the faceting variable
  sample_sort_column: "Date"      # column containing the sorting variable
  sample_filter_column: "Include" # use this column for filtering
  sample_filter_value: "Yes"      # only plot samples which have this value in the "filter" column
---

## Parameters

This analysis uses the following parameters that can be modified when rendering:

```{r display-params}
# ensure taxonomic level is lowercase
params$taxonomic_level <- tolower(params$taxonomic_level)
params$bubble_plot_taxonomic_level_color <- tolower(params$bubble_plot_taxonomic_level_color)

# check for spelling errors
tax_ranks_ordered <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")

if (!params$taxonomic_level %in% tax_ranks_ordered) {
  stop("Invalid taxonomic rank for aggregation!")
}
if (!params$bubble_plot_taxonomic_level_color %in% tax_ranks_ordered) {
  stop("Invalid taxonomic rank for bubble plot coloring!")
}

# display current parameter settings
params
```

The optional sample data (e.g., `sample_data.xlsx`) file has sample name in the first column, and sample "type" and "sorting" columns using the names set in the parameters section.

## Setup

Load required packages for the analysis:

```{r setup}
# load required packages (install if necessary)
install_if_needed <- function(pkg, bioc = FALSE) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (bioc) {
      if (!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
      BiocManager::install(pkg)
    } else {
      install.packages(pkg)
    }
  }
  library(pkg, character.only = TRUE)
}

# install and load packages
install_if_needed('phyloseq', bioc=TRUE)
install_if_needed('ggplot2')
install_if_needed('dplyr')
install_if_needed('purrr')
install_if_needed('Polychrome')
install_if_needed('tibble')
install_if_needed('readxl')
install_if_needed('stringr')

cat("All packages loaded successfully.\n")
```

## Load Data

Loading data from the specified directory:

```{r load-data}
input_dir <- params$input_dir

seqtab.nochim <- readRDS(file.path(input_dir, 'seqtab.nochim.rds'))
taxa <- readRDS(file.path(input_dir, 'taxa.rds'))
sample.names <- rownames(seqtab.nochim)

# load the sample data if requested and check formatting
has_sample_data <- length(params$sample_data_file)
if (has_sample_data) {
  sample_data <- read_excel(file.path(input_dir, params$sample_data_file))

  # take the first column of the sample_data file and put it into rownames instead
  sample_data <- column_to_rownames(sample_data, var = colnames(sample_data)[1])

  # ensure sample order matches seqtab.nochim
  if (!all(rownames(sample_data) %in% sample.names)) {
    warning("Some sample names in sample data don't match the sequence table.")
  }

  # match the sample data to the sequence table order
  prev_sample_data <- sample_data
  sample_data <- sample_data[match(sample.names, rownames(sample_data)), ]
  
  if (length(prev_sample_data[,1]) != length(sample_data[,1])) {
    stop("Some samples in the sample data file are not present in the sequence table. The sample data file and the sequence data must match.")
  }

  # apply filtering if sample_filter_column is specified
  if (!is.null(params$sample_filter_column) && params$sample_filter_column %in% colnames(sample_data)) {
    cat(sprintf("* Filtering samples based on '%s' column with value '%s'...\n", 
                params$sample_filter_column, params$sample_filter_value))
    
    # get samples to keep
    samples_to_keep <- rownames(sample_data)[sample_data[[params$sample_filter_column]] == params$sample_filter_value]
    
    # filter sequence table
    seqtab.nochim <- seqtab.nochim[rownames(seqtab.nochim) %in% samples_to_keep, ]
    
    # update sample names
    sample.names <- rownames(seqtab.nochim)
    
    cat(sprintf("   Keeping %d samples after filtering.\n", length(samples_to_keep)))
  }

  if (!params$sample_type_column %in% colnames(sample_data) || 
      !params$sample_sort_column %in% colnames(sample_data)) {
    stop(paste("Required columns not found in sample data file.",
               "Looking for:", params$sample_type_column, 
               "and", params$sample_sort_column))
  }
  
  sample_data  # display sample data if we have it
}
```

## Create the Phyloseq Object

Here we'll create the `phyloseq` object containing the OTU/ASV table and taxonomic data needed for plotting. If we have additional sample data from the `sample_data.xlsx` file, it goes into this object as well.

```{r phyloseq-object}
if (has_sample_data) {
  ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
                 sample_data(sample_data), tax_table(taxa))
} else {
  ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), tax_table(taxa))
}
ps
```

## Filter

In this cell, we prepare the data. Counts are converted to relative abundances, low-abundance taxa are grouped into the "Minor taxa" group, and abundant taxa are agglomerated (clustering several ASVs which have the same classification). We also clean up the names for plotting later.

```{r filter-aggregate-taxa}
# set the taxonomic level and abundance threshold for analysis
tax_level <- params$taxonomic_level
min_abund_threshold <- params$min_abundance

cat(paste0("\n--- Filtering and Aggregation at ", tax_level, " level (Threshold: ", min_abund_threshold * 100, "%) ---\n"))

### transform counts to relative abundance
cat("* Calculating relative abundance for all taxa...\n")
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

### identify abundant taxa
#   - abundance threshold is max per sample, i.e.,
#     it's enough for a taxon to be above the threshold in a single sample
cat(paste0("* Identifying taxa with max relative abundance >= ", 
           min_abund_threshold * 100, "% in any sample...\n"))
abund_mat <- as(otu_table(ps_rel), "matrix") # convert otu table to matrix
if (ncol(abund_mat) == 0) stop("Abundance matrix has no taxa after initial steps.")

# handle case where there's only one sample
if (nrow(abund_mat) == 1) {
  max_abundance_per_taxon <- abund_mat[1, ]
} else {
  max_abundance_per_taxon <- apply(abund_mat, 2, max) # max abundance per taxon (column)
}

# extract abundant OTUs
abundant_taxa_otus <- names(max_abundance_per_taxon[max_abundance_per_taxon >= min_abund_threshold])
cat(sprintf("   Found %d OTUs/ASVs meeting the threshold out of %d total.\n",
            length(abundant_taxa_otus), ntaxa(ps_rel)))

### pruning
cat("* Pruning phyloseq object to keep only these abundant taxa...\n")
# handle case where no taxa meet the threshold
if (length(abundant_taxa_otus) == 0) {
  warning("No taxa met the abundance threshold. Plot will only show 'Minor taxa'.")
  ps_rel_abundant_only <- NULL
} else {
  ps_rel_abundant_only <- prune_taxa(abundant_taxa_otus, ps_rel)
  cat(sprintf("   Pruned object contains %d taxa.\n", ntaxa(ps_rel_abundant_only)))
}


### agglomeration
#   here we join similar ASVs for a cleaner plot
cat(sprintf("* Agglomerating abundant taxa at the '%s' level...\n", tax_level))
if (params$agglomerate && !is.null(ps_rel_abundant_only)) {
  # ensure the taxonomic level exists in the analysis object
  if (!tax_level %in% rank_names(ps_rel_abundant_only)) {
      stop(paste("Taxonomic rank", tax_level, "not found in taxonomy table."))
  }

  # Perform agglomeration - NArm=FALSE keeps taxa with NA at the target rank
  ps_merged_abundant <- tax_glom(ps_rel_abundant_only, taxrank = tax_level, NArm = FALSE)
  cat(sprintf("   Agglomerated into %d groups at the %s level.\n", ntaxa(ps_merged_abundant), tax_level))
} else if (!params$agglomerate) {
  cat("   Skipping agglomeration as params$agglomerate == false.")
  ps_merged_abundant <- ps_rel_abundant_only
} else {
  ps_merged_abundant <- NULL # No abundant taxa to agglomerate
  cat("   Skipping agglomeration as no taxa met the threshold.\n")
}

### cleaning up display names
cat("* Generating unique display names for agglomerated taxa groups...\n")
if (!is.null(ps_merged_abundant)) {
    tax_table_merged_mat <- as(tax_table(ps_merged_abundant), "matrix")
    tax_table_merged <- as_tibble(tax_table_merged_mat, rownames = "OTU_ID") # keep original OTU IDs from tax_glom

    # filter to ranks actually present in the data
    tax_ranks_present <- intersect(tax_ranks_ordered, colnames(tax_table_merged))
    
    # clean names - clean_name(name)
    #
    # this function converts the _123 suffix to [123], if desired.
    clean_name <- function(name) {
      if (params$clean_names) {
        str_replace(name, "_(\\d+)$", " [\\1]")
      } else {
        name
      }
    }

    # initialize counter for unidentified taxa before calling get_display_name
    unidentified_counter <- 0
    
    # create display names - get_display_name(tax_row, target_level, all_ranks)
    # 
    # this function looks for the lowest rank with an identifier. 
    # i.e., for a species without an identification at the specified level,
    # (for example "genus"), the resulting name might be "Bacillota (phylum)" or
    # "Planococcaceae (family)".
    get_display_name <- function(tax_row, target_level, all_ranks) {
        target_val <- tax_row[[target_level]]
        target_rank_idx <- which(all_ranks == target_level)

        # check whether we have an identification at the specified rank
        if (is.na(target_val) || target_val == "" || grepl('^Incertae Sedis', target_val)) {
            # find the closest higher rank that IS classified
            higher_rank_val <- NA
            higher_rank_name <- NA
            if (target_rank_idx > 1) {
                for (i in (target_rank_idx - 1):1) {
                    rank_to_check <- all_ranks[i]
                    val_to_check <- tax_row[[rank_to_check]]
                    if (!is.na(val_to_check) && val_to_check != "" && !grepl('^Incertae Sedis', val_to_check)) {
                        higher_rank_val <- val_to_check
                        higher_rank_name <- rank_to_check
                        break  # found it
                    }
                }
            }
            
            # If we couldn't find any identification at any rank
            if (is.na(higher_rank_val) || higher_rank_val == "") {
                # Increment counter and return numbered "Unidentified"
                unidentified_counter <<- unidentified_counter + 1
                return(paste0("Unidentified [", unidentified_counter, "]"))
            } else {
                # construct name: "HigherRankName (rank)"
                return(paste0(clean_name(higher_rank_val), " (", higher_rank_name, ")"))
            }
        } else {
            # target level is classified, just return its value
            return(clean_name(target_val))
        }
    }

    tax_table_merged$display_name <- apply(tax_table_merged, 1, get_display_name, target_level = tax_level, all_ranks = tax_ranks_present)

    # if agglomeration was skipped, ASVs with the same classification might have the same display name.
    # we need to make them unique before assigning them as taxa_names.
    if (!params$agglomerate) {
        cat("   Agglomeration skipped, ensuring unique display names for individual ASVs...\n")
        # check for duplicate display names generated
        name_counts <- table(tax_table_merged$display_name)
        duplicated_names <- names(name_counts[name_counts > 1])

        if (length(duplicated_names) > 0) {
            cat(sprintf("      Found %d display names shared by multiple ASVs. Appending suffix for uniqueness.\n", length(duplicated_names)))
            # iterate through rows and append OTU_ID if the display_name is duplicated
            tax_table_merged <- tax_table_merged %>%
              group_by(display_name) %>%
              # add a unique suffix (_1, _2, etc.) only if the display name is duplicated
              mutate(
                n_in_group = n(),
                unique_suffix = if_else(n_in_group > 1, paste0("_", row_number()), "")
              ) %>%
              ungroup() %>%
              mutate(display_name = paste0(display_name, unique_suffix)) %>%
              select(-n_in_group, -unique_suffix)
            cat("      Appended OTU IDs to duplicated names.\n")
        } else {
            cat("      All generated display names are already unique.\n")
        }
    }

    # create a mapping from original OTU_ID to new (potentially modified) display name
    name_map <- setNames(tax_table_merged$display_name, tax_table_merged$OTU_ID)

    # apply the new names to the phyloseq object
    taxa_names(ps_merged_abundant) <- name_map[taxa_names(ps_merged_abundant)]
    cat(sprintf("   Assigned %d unique display names.\n", length(taxa_names(ps_merged_abundant))))
} else {
    cat("   Skipping display name generation.\n")
}


### prepare the data for plotting
cat("* Melting agglomerated abundant taxa data (if any)...\n")
if (!is.null(ps_merged_abundant)) {
    ps_merged_abundant_df <- psmelt(ps_merged_abundant)
    # ensure Abundance is numeric
    ps_merged_abundant_df$Abundance <- as.numeric(ps_merged_abundant_df$Abundance)

    # add taxonomic rank for bubble plot coloring
    bubble_color_rank <- params$bubble_plot_taxonomic_level_color
    if (bubble_color_rank %in% colnames(ps_merged_abundant_df)) {
        # rename the column to avoid conflicts and make it generic
        bubble_color_col_name <- paste0("bubble_color_", bubble_color_rank)
        ps_merged_abundant_df[[bubble_color_col_name]] <- ps_merged_abundant_df[[bubble_color_rank]]
        cat(sprintf("   Added column '%s' for bubble plot coloring based on rank '%s'.\n", bubble_color_col_name, bubble_color_rank))
    } else {
        warning(paste(
          "Bubble plot coloring rank '", bubble_color_rank, 
          "' not found in taxonomy data. Coloring will default to OTU name."))
        bubble_color_col_name <- "OTU" # fallback to OTU if rank not found
    }

    cat(sprintf("   Melted data has %d rows.\n", nrow(ps_merged_abundant_df)))
} else {
    # create an empty dataframe with expected columns if no abundant taxa
    expected_cols <- c("Sample", "OTU", "Abundance")
    bubble_color_rank <- params$bubble_plot_taxonomic_level_color
    bubble_color_col_name <- paste0("bubble_color_", bubble_color_rank) # define even if empty
    expected_cols <- c(expected_cols, bubble_color_col_name) # add color column name

    if (has_sample_data) {
        # add sample data columns that exist in the original sample_data
        sample_data_cols <- colnames(sample_data(ps))
        expected_cols <- c(expected_cols, intersect(c(params$sample_type_column, params$sample_sort_column), sample_data_cols))
    }
    ps_merged_abundant_df <- data.frame(matrix(ncol = length(expected_cols), nrow = 0))
    colnames(ps_merged_abundant_df) <- expected_cols
    # ensure correct types
    ps_merged_abundant_df <- ps_merged_abundant_df %>%
        mutate(Sample = as.character(Sample),
               OTU = as.character(OTU),
               Abundance = as.numeric(Abundance),
               # use !! operator to use the value of the variable as the column name
               !!bubble_color_col_name := as.character(.data[[bubble_color_col_name]]))

    if (has_sample_data) {
        if(params$sample_type_column %in% colnames(ps_merged_abundant_df)) {
          # sorting column type might vary, handle later during plotting
          ps_merged_abundant_df[[params$sample_type_column]] <- factor()
        }
    }
    cat("   No abundant taxa data to melt.\n")
}


### add the "minor taxa" group
cat("* Calculating abundance of minor taxa (sum of all taxa below threshold)...\n")
# sum the abundances of the *agglomerated abundant* taxa for each sample
if (nrow(ps_merged_abundant_df) > 0) {
    total_abundant_per_sample <- ps_merged_abundant_df %>%
      group_by(Sample) %>%
      summarize(total_abundant = sum(Abundance, na.rm = TRUE), .groups = 'drop')
} else {
    # if no abundant taxa, their sum is 0 for all samples
    all_samples <- sample_names(ps)
    total_abundant_per_sample <- tibble(Sample = all_samples, total_abundant = 0.0)
}

# create the "minor taxa" group
minor_taxa_label <- paste0("Minor taxa (<", min_abund_threshold * 100, "%)")
minor_taxa_df <- total_abundant_per_sample %>%
  mutate(OTU = minor_taxa_label, Abundance = 1.0 - total_abundant) %>%
  select(Sample, OTU, Abundance)

# add the bubble color column to minor taxa df
minor_taxa_color_group_label <- "Minor Taxa Group" # specific label for coloring
minor_taxa_df[[bubble_color_col_name]] <- minor_taxa_color_group_label


# add sample data columns to minor_taxa_df if sample_data exists
if (has_sample_data) {
  sample_data_df_orig <- as_tibble(sample_data(ps), rownames = "Sample")
  # keep only relevant sample data columns
  sample_data_to_join <- sample_data_df_orig %>% 
    select(Sample, any_of(c(params$sample_type_column, params$sample_sort_column)))
  minor_taxa_df <- minor_taxa_df %>%
    left_join(sample_data_to_join, by = "Sample")
}


### combine abundant and minor taxa dfs
cat("* Combining abundant and minor taxa into a single dataframe for plotting...\n")
# ensure both dataframes have the same essential columns before binding
common_cols <- c("Sample", "OTU", "Abundance", bubble_color_col_name)
if (has_sample_data) {
    # get column names that actually exist in the original sample data
    existing_sample_cols <- intersect(c(params$sample_type_column, params$sample_sort_column), colnames(sample_data(ps)))
    common_cols <- unique(c(common_cols, existing_sample_cols))
}

# select common columns, ensuring they exist in both dataframes
# need to handle case where ps_merged_abundant_df might not have the bubble color column if fallback occurred
if (!bubble_color_col_name %in% colnames(ps_merged_abundant_df)) {
    # if the color column doesn't exist (e.g., fallback to OTU), create it from OTU
    ps_merged_abundant_df[[bubble_color_col_name]] <- ps_merged_abundant_df[["OTU"]]
}
ps_merged_abundant_df_select <- ps_merged_abundant_df %>% select(any_of(common_cols))
minor_taxa_df_select <- minor_taxa_df %>% select(any_of(common_cols))

# combine
combined_df <- bind_rows(ps_merged_abundant_df_select, minor_taxa_df_select)

# ensure the bubble color column is treated as a factor for coloring
combined_df[[bubble_color_col_name]] <- factor(combined_df[[bubble_color_col_name]])


cat(sprintf("   Final combined dataframe has %d rows.\n", nrow(combined_df)))


cat("--- Filtering and Aggregation Complete ---\n")
```

## Plot

```{r plot-relative-abundance}
cat("\n--- Plotting Relative Abundance ---\n")

# ensure combined_df exists and has data
if (!exists("combined_df") || nrow(combined_df) == 0) {
  stop("The combined dataframe for plotting is empty. Check filtering steps.")
}

### generate the palettes
#   we're using Polychrome::createPalette() to make a good palette for this type of data.
#   see https://cran.r-project.org/web/packages/Polychrome/vignettes/creatingPalettes.html
if (!is.null(ps_merged_abundant) && nrow(ps_merged_abundant_df) > 0) {
    # sort abundant taxa by overall mean abundance for consistent ordering
    mean_abund_sorted_taxa <- ps_merged_abundant_df %>%
        filter(OTU != minor_taxa_label) %>% # probably not needed
        group_by(OTU) %>%
        summarize(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = 'drop') %>%
        arrange(desc(mean_abundance)) %>%
        pull(OTU)
    abundant_taxa_names <- mean_abund_sorted_taxa
} else {
    abundant_taxa_names <- character(0) # no abundant taxa
}
ntaxa_abundant <- length(abundant_taxa_names)
cat(sprintf("* Preparing color palette for %d abundant taxa groups (barplot).\n", ntaxa_abundant))


# Polychrome claims it can generate at least 38 distinguishable colors
# - if you need more then you're probably plotting your data wrong!
if (ntaxa_abundant > 38) {
  warning(paste("High number of abundant taxa groups (", ntaxa_abundant,
                "). Colors in the plot may be difficult to distinguish.",
                "Consider using a higher taxonomic level or abundance threshold."))
}

set.seed(924)  # for reproducibility
if (ntaxa_abundant > 0) {
  # use a broad set of seed colors
  # the seed colors may not matter much, but they do influence ordering and can probably
  # give your palettes a different style
  # there are online palette generators you can use for experimenting with seed colors
  seed_cols <- c("#e41a1c", 
                 "#377eb8", 
                 "#4daf4a", 
                 "#984ea3", 
                 "#ff7f00", 
                 "#ffff33", 
                 "#a65628", 
                 "#f781bf", 
                 "#999999")
  abundant_colors_stack <- Polychrome::createPalette(ntaxa_abundant, seed_cols)
  names(abundant_colors_stack) <- abundant_taxa_names
  cat(sprintf("   Generated palette with %d distinct colors for stacked barplot.\n", length(abundant_colors_stack)))
} else {
  abundant_colors_stack <- character(0) # empty palette if no abundant taxa
  cat("   No abundant taxa, stacked barplot palette will only contain minor taxa color.\n")
}


# add minor taxa color and create final palette (for stacked barplot)
minor_taxa_color <- c("grey45")
names(minor_taxa_color) <- minor_taxa_label
final_colors_stack <- c(abundant_colors_stack, minor_taxa_color)

# create palette for bubble plot based on bubble_color_col_name
bubble_color_rank <- params$bubble_plot_taxonomic_level_color
bubble_color_col_name <- paste0("bubble_color_", bubble_color_rank)

bubble_color_groups <- levels(combined_df[[bubble_color_col_name]]) # get all unique groups including minor
# remove the minor taxa group temporarily to generate palette for actual taxa ranks
bubble_color_groups_taxa_only <- setdiff(bubble_color_groups, minor_taxa_color_group_label)
n_bubble_color_groups_taxa <- length(bubble_color_groups_taxa_only)

cat(sprintf("* Preparing color palette for %d groups based on rank '%s' (bubble plot).\n",
            length(bubble_color_groups), bubble_color_rank))

if (n_bubble_color_groups_taxa > 0) {
  if (n_bubble_color_groups_taxa > 38) {
    # this is unlikely to happen but might as well check
    warning(paste("High number of coloring groups (", n_bubble_color_groups_taxa,
                  ") for bubble plot based on rank '", bubble_color_rank,
                  "'. Colors may be hard to distinguish."))
  }
  # i like a pastel look for the bubbles
  seed_cols_bubble <- c("#a6cee3", 
                        "#b2df8a", 
                        "#fb9a99", 
                        "#fdbf6f", 
                        "#cab2d6", 
                        "#ffff99")
  bubble_colors_taxa <- Polychrome::createPalette(n_bubble_color_groups_taxa, seed_cols_bubble)
  names(bubble_colors_taxa) <- bubble_color_groups_taxa_only
  cat(sprintf("   Generated palette with %d distinct colors for taxa ranks.\n", length(bubble_colors_taxa)))
} else {
    bubble_colors_taxa <- character(0)
    cat("   No specific taxa ranks found for coloring, only minor group.\n")
}

# add minor taxa group color (use the same grey as barplot)
bubble_minor_color <- c("grey45")
names(bubble_minor_color) <- minor_taxa_color_group_label
final_colors_bubble <- c(bubble_colors_taxa, bubble_minor_color)

# ensure the order matches the factor levels for the legend
final_colors_bubble <- final_colors_bubble[bubble_color_groups]


### prepare factors for plotting
#   this might need some tweaking for your specific data
cat("* Setting factor levels for plotting order...\n")
# order taxa factors for stacking: abundant taxa first (sorted by mean abundance), then minor taxa
stacking_order_levels <- c(abundant_taxa_names, minor_taxa_label)
# ensure all OTU values in combined_df are in the levels
combined_df$OTU <- factor(combined_df$OTU, levels = stacking_order_levels)

# define desired *legend* order: abundant taxa (sorted) first, then minor taxa
legend_order_breaks <- c(abundant_taxa_names, minor_taxa_label)

# order samples based on sample data
has_facet_column <- FALSE  # default
has_sort_column <- FALSE   # default

if (has_sample_data) {
  # check if the specified columns actually exist in the combined_df (added during merge)
  has_sort_column <- params$sample_sort_column %in% colnames(combined_df)
  has_facet_column <- params$sample_type_column %in% colnames(combined_df)

  if (has_sort_column) {
    # get the sample order based on the sorting column
    # need the original sample_data_df_orig for this
    if (!exists("sample_data_df_orig")) {
        sample_data_df_orig <- as_tibble(sample_data(ps), rownames = "Sample")
    }

    # Handle different data types for sorting (numeric, date, character)
    sort_col_data <- sample_data_df_orig[[params$sample_sort_column]]
    if (inherits(sort_col_data, "Date") || inherits(sort_col_data, "POSIXt")) {
      # sort column is date
      sample_order <- sample_data_df_orig %>% 
        arrange(.data[[params$sample_sort_column]]) %>% 
        pull(Sample)
    } else if (is.numeric(sort_col_data)) {
      # sort coumn is numeric
       sample_order <- sample_data_df_orig %>% 
         arrange(.data[[params$sample_sort_column]]) %>% 
         pull(Sample)
    } else {
      # sort column is character - assume factor and sort alphabetically
      sample_order <- sample_data_df_orig %>% 
        arrange(as.character(.data[[params$sample_sort_column]])) %>% 
        pull(Sample)
    }
    
    # filter sample_order to only include samples present in combined_df
    sample_order <- sample_order[sample_order %in% unique(combined_df$Sample)]
    combined_df$Sample <- factor(combined_df$Sample, levels = sample_order)
    cat(sprintf("   Ordered samples based on '%s'.\n", params$sample_sort_column))
  } else {
    warning(paste("Sample sort column '", params$sample_sort_column, 
                  "' not found in combined data. Samples will be ordered alphabetically."))
    combined_df$Sample <- factor(combined_df$Sample, 
                                 levels = sort(unique(combined_df$Sample)))
  }

  # ensure facet column is a factor if it exists
  if (has_facet_column) {
    combined_df[[params$sample_type_column]] <- factor(combined_df[[params$sample_type_column]])
    cat(sprintf("   Set '%s' as factor for faceting.\n", params$sample_type_column))
  } else {
    warning(paste("Sample type column '", params$sample_type_column, "' not found in combined data. Faceting will be skipped."))
  }
} else {
  # If no sample data, order samples alphabetically
  combined_df$Sample <- factor(combined_df$Sample, levels = sort(unique(combined_df$Sample)))
  cat("   Ordered samples alphabetically (no sample data for sorting/faceting).\n")
}


### make the barplot
cat("* Creating the stacked barplot...\n")
barplot <- ggplot(combined_df, aes(x = Sample, y = Abundance, fill = OTU)) +
  geom_bar(stat = "identity", position = "stack", color = "black", size = 0.15) +
  scale_fill_manual(
    values = final_colors_stack,
    breaks = legend_order_breaks, # explicitly set legend order
    labels = function(l) ifelse(l == minor_taxa_label, 
                                paste0("Minor (<", params$min_abundance * 100, "%)"), 
                                l), # rewrite minor label to include filtering level
    drop = FALSE # keep all levels in the legend even if abundance is zero in plotted subset
  ) +
  theme_minimal(base_size = 11) +
  labs(
    title = if (params$plot_titles) { 
      paste("Relative Abundance by Sample (Aggregated at", tax_level, "level)")
    },
    subtitle = if (params$plot_titles) { 
      paste("Taxa shown if >", params$min_abundance * 100, 
            "% relative abundance in any sample. Others grouped into 'Minor taxa'.")
      },
    x = if(has_sort_column) {
        paste("Sample (ordered by", params$sample_sort_column, ")") 
      } else { "Sample" },
    y = "Relative Abundance",
    fill = str_to_sentence(tax_level) # capitalize first letter in taxonomic rank
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1, size = rel(0.9)),
    legend.position = "right",
    legend.key.size = unit(0.8, "lines"),
    legend.title = element_text(size = rel(0.9), face = "bold"),
    legend.text = element_text(size = rel(0.85)),
    strip.text = element_text(face = "bold", size=rel(1.0)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face="bold", size=rel(1.2)),
    plot.subtitle = element_text(size=rel(0.9))
  )

# add faceting if sample type column is specified and exists
if (has_facet_column && params$sample_type_column != "") {
  cat(sprintf("   Adding faceting by '%s' to barplot.\n", params$sample_type_column))
  # create formula for facet_grid
  facet_formula <- as.formula(paste("~", params$sample_type_column))
  barplot <- barplot + 
    facet_grid(cols = vars(.data[[params$sample_type_column]]),
               scales = "free_x", # allow x-axis to vary between facets
               space = "free_x") # allow space allocated to vary
}

# display the barplot
print(barplot)
ggsave(file.path(input_dir, paste0("relative_abundance_barplot_", params$taxonomic_level, ".pdf")), plot=barplot, width=25, height=15, units="cm")
cat("   Barplot saved.\n")


### make the bubble plot
cat("* Creating the bubble plot...\n")

# filter out zero-abundance points for bubble plot clarity
combined_df_bubble <- combined_df %>% filter(Abundance > 0)

# ensure OTU is factor with correct levels for bubble plot y-axis
combined_df_bubble$OTU <- factor(combined_df_bubble$OTU, levels = rev(stacking_order_levels)) # most abundant (overall) taxa on top

# ensure the bubble color column is factor with correct levels
combined_df_bubble[[bubble_color_col_name]] <- factor(combined_df_bubble[[bubble_color_col_name]], levels = bubble_color_groups)

# reorder bubble_color_groups to keep alphabetical order but put minor taxa last
bubble_color_groups_taxa_only <- setdiff(bubble_color_groups, minor_taxa_color_group_label)
bubble_color_groups_taxa_only <- sort(bubble_color_groups_taxa_only)
bubble_color_groups_reordered <- c(bubble_color_groups_taxa_only, minor_taxa_color_group_label)

# update the factor levels with the new order
combined_df_bubble[[bubble_color_col_name]] <- factor(combined_df_bubble[[bubble_color_col_name]], levels = bubble_color_groups_reordered)

# reorder the colors to match the new order
final_colors_bubble <- final_colors_bubble[bubble_color_groups]
final_colors_bubble_reordered <- final_colors_bubble[bubble_color_groups_reordered]


bubbleplot <- ggplot(combined_df_bubble, aes(x=Sample, y=OTU, size=Abundance, fill=.data[[bubble_color_col_name]])) + # use the dynamic color column name
  geom_point(alpha = 0.8, color="black", shape=21) + # shape=21 allows black outline
  scale_fill_manual(
    name = str_to_sentence(params$bubble_plot_taxonomic_level_color),
    values = final_colors_bubble_reordered,
    breaks = bubble_color_groups_reordered, # use the reordered groups from the color rank column
    labels = function(l) 
      ifelse(l == minor_taxa_color_group_label, 
             paste0("Minor/Unknown"), 
             l),
    guide = "legend",
    drop = FALSE
  ) +
  scale_size_area(
    max_size = 10,
    name = "Relative\nAbundance (%)",
    labels = scales::percent_format(accuracy = 1)
  ) +
  theme_minimal(base_size = 11) +
  labs(
    title = if (params$plot_titles) {
      paste("Relative Abundance Bubble Plot (Aggregated at", tax_level, "level)")
      },
    subtitle = if (params$plot_titles) {
      paste("Bubble size indicates relative abundance (> 0%). Color indicates",
                     params$bubble_plot_taxonomic_level_color, "rank.")
      },
    x = if(has_sort_column) { 
        paste("Sample (ordered by", params$sample_sort_column, ")")
      } else { "Sample" },
    y = str_to_sentence(tax_level)
    ) +
    theme(
      axis.text.x = element_text(angle = 60, hjust = 1, size = rel(0.9)),
      axis.text.y = element_text(size = rel(0.9)),
      legend.position = "right",
      legend.key.size = unit(0.8, "lines"),
      legend.title = element_text(size = rel(0.9), face = "bold"),
      legend.text = element_text(size = rel(0.85)),
      strip.text = element_text(face = "bold", size=rel(1.0)),
      panel.grid.major = element_line(color = "grey90", size = 0.2),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face="bold", size=rel(1.2)),
      plot.subtitle = element_text(size=rel(0.9))
    )

# add faceting if sample type column is specified and exists
if (has_facet_column && params$sample_type_column != "") {
  cat(sprintf("   Adding faceting by '%s' to bubble plot.\n",
              params$sample_type_column))
  facet_formula <- as.formula(paste("~", params$sample_type_column))
  bubbleplot <- bubbleplot + 
    facet_grid(cols = vars(.data[[params$sample_type_column]]),
               scales = "free_x",
               space = "free_x")
}

# display the bubble plot
print(bubbleplot)
ggsave(file.path(input_dir, 
                 paste0("relative_abundance_bubbleplot_", 
                        params$taxonomic_level, "_color_",
                        params$bubble_plot_taxonomic_level_color, ".pdf")), 
       plot=bubbleplot, 
       width=25, height=15 + ntaxa_abundant*0.2, 
       units="cm", limitsize = FALSE)

cat("   Bubble plot saved.\n")
cat("--- Plotting Complete ---\n")
```

End of script.
